---
title: "Spatial ChIP (ChIPSP) as a new Bioinformatics Tool to characterize spatial gene regulation"
author: "Tianyi Zhou"
package: ChIPSP
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ChIPSP Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

# Introduction

`ChIPSP` is an R package for integrating ChIP-seq transcription factor binding
with Hi-C chromatin interaction data to identify spatially linked regulatory
regions.

The ChIP-SP workflow combines information from:

- ChIP-seq peak position and signal strength
- Hi-C chromatin loop coordinates
- Hi-C loop confidence
- gene annotation
- pathway and functional enrichment

The current workflow consists of five major steps:

1. Optional removal of chromosome X and chromosome Y interactions
2. Merging Hi-C loop files across replicates and/or resolutions
3. ChIP-SP spatial integration and ranking
4. Gene annotation
5. Enrichr functional enrichment analysis

The major exported functions are:

```{r functions}
library(ChIPSP)

getNamespaceExports("ChIPSP")
```

The primary workflow functions are:

```text
removeXYChromosomes()
mergeHiCLoops()
readChIPFile()
chipSPLink()
ChIPSPannotation()
ChIPSPenrichment()
```

# Conceptual Framework

Hi-C identifies genomic regions that interact in three-dimensional chromatin
space. However, a Hi-C loop anchor generally represents a genomic interval
rather than an exact nucleotide-level regulatory element.

ChIP-SP integrates ChIP-seq peaks with these Hi-C loop anchors.

If a ChIP-seq peak overlaps one anchor of a Hi-C loop, the interacting partner
anchor is identified as a spatially linked regulatory region.

Conceptually:

```text
ChIP-seq peak
      |
      v
Hi-C anchor 1
      |
      | chromatin interaction
      v
Hi-C anchor 2
      |
      v
Spatially linked regulatory region
```

The same procedure is performed in both directions:

```text
ChIP overlap with BIN1 -> project to BIN2
ChIP overlap with BIN2 -> project to BIN1
```

ChIP-SP then ranks spatially linked regions using information from both
ChIP-seq and Hi-C.

The ranking incorporates:

- ChIP-seq signal strength (`pileup`)
- Hi-C loop confidence (`FDR`)

The current ChIP-SP score is based on normalized ChIP-seq signal and normalized
Hi-C FDR:

```text
score = pileup_norm - fdr_norm
```

Higher ChIP-seq signal and lower Hi-C FDR therefore favor a higher ChIP-SP
ranking.

# Installation

The development version of `ChIPSP` can be installed from GitHub.

If a previous version is already installed, it can first be removed:

```{r install, eval=FALSE}
remove.packages("ChIPSP")
```

Install the current version from the `main` branch:

```{r install-github, eval=FALSE}
install.packages("remotes")

remotes::install_github(
  "Lattesnow/ChIPSP@main",
  force = TRUE,
  upgrade = "never"
)
```

Load the package:

```{r load-package, eval=FALSE}
library(ChIPSP)
```

# Input Data

A typical ChIP-SP analysis requires:

1. one or more Hi-C loop files
2. a ChIP-seq peak file

Input files may be placed in the current working directory or supplied using
full file paths.

## Hi-C Input

Hi-C files should contain the two genomic loop anchors together with an
interaction confidence value.

The merged Hi-C table used by `chipSPLink()` contains fields corresponding to:

```text
BIN1_CHR
BIN1_START
BIN1_END
BIN2_CHR
BIN2_START
BIN2_END
FDR
```

Multiple Hi-C files may represent:

- biological replicates
- different loop-calling resolutions
- different compatible Hi-C datasets

These files can be combined using `mergeHiCLoops()`.

## ChIP-seq Input

ChIP-seq peak files may be provided in commonly used formats including:

```text
BED
CSV
TSV
TXT
TAB
XLS
XLSX
```

The ChIP-SP analysis requires genomic coordinates and ChIP-seq signal
information.

The ChIP-seq signal used in ChIP-SP ranking is represented by:

```text
pileup
```

The ChIP-seq and Hi-C data should use the same reference genome.

For example:

```text
hg19 ChIP-seq + hg19 Hi-C
hg38 ChIP-seq + hg38 Hi-C
mm9 ChIP-seq + mm9 Hi-C
mm10 ChIP-seq + mm10 Hi-C
```

Genome builds should not be mixed within the same analysis.

# Step 1: Remove Chromosome X and Y

`removeXYChromosomes()` optionally removes interactions involving chromosome X
or chromosome Y.

When no file names are provided, the function can search the current directory
for compatible Hi-C files.

```{r remove-xy, eval=FALSE}
hic_files_clean <- removeXYChromosomes(
  path = getwd()
)
```

The returned object contains the paths of the filtered Hi-C files.

These paths can be passed directly to `mergeHiCLoops()`.

```{r remove-xy-output, eval=FALSE}
hic_files_clean
```

This step is optional.

If chromosome X or chromosome Y interactions are relevant to the biological
question, the original Hi-C files can instead be passed directly to the next
step.

# Step 2: Merge Hi-C Loop Files

`mergeHiCLoops()` combines Hi-C loop files into one unified loop table.

For example:

```{r merge-hic, eval=FALSE}
hic_df <- mergeHiCLoops(
  hic_files_clean
)
```

The resulting object is a `data.frame` containing the merged Hi-C loops.

```{r inspect-hic, eval=FALSE}
head(hic_df)
dim(hic_df)
```

The merged table can then be passed directly to `chipSPLink()`.

# Step 3: ChIP-SP Spatial Integration

`chipSPLink()` performs the central ChIP-SP spatial integration.

A minimal call is:

```{r chipsp, eval=FALSE}
chipsp_results <- chipSPLink(
  hic_df = hic_df
)
```

When `chip_file` is not explicitly supplied, `chipSPLink()` uses
`readChIPFile()` to identify a compatible ChIP-seq file in the working
directory.

For example, a file named:

```text
r1881_hg19_test_peaks_ChIP.bed
```

can be automatically detected.

The ChIP file can also be specified explicitly:

```{r chipsp-explicit, eval=FALSE}
chipsp_results <- chipSPLink(
  chip_file = "r1881_hg19_test_peaks_ChIP.bed",
  hic_df = hic_df
)
```

By default, the results are written to:

```text
ChIPSP_results.csv
```

The returned object is also available directly in R:

```{r inspect-chipsp, eval=FALSE}
head(chipsp_results)
dim(chipsp_results)
```

Typical ChIP-SP output columns include:

```text
chr
start
end
pileup
FDR
source_anchor
pileup_norm
fdr_norm
score
rank
```

`source_anchor` indicates whether the spatial projection originated from the
first or second Hi-C loop anchor.

Duplicate projected rows generated within the same projection process are
removed. However, an identical spatial interaction independently generated
from the two different anchor-projection processes is retained as an
independent observation.

# Step 4: Gene Annotation

`ChIPSPannotation()` annotates both conventional ChIP-seq peaks and ChIP-SP
spatially linked regions to genes.

The function supports human and mouse reference genomes.

## Supported Species and Genome Builds

The currently supported combinations are:

| Species | `species` | `ref_genome` |
|:--------|:----------|:-------------|
| Human | `"human"` | `"hg19"` |
| Human | `"human"` | `"hg38"` |
| Mouse | `"mouse"` | `"mm9"` |
| Mouse | `"mouse"` | `"mm10"` |

The species and genome build must correspond.

For example:

```text
human + hg19   valid
human + hg38   valid
mouse + mm9    valid
mouse + mm10   valid

human + mm10   invalid
mouse + hg38   invalid
```

## Human hg19

For human data aligned to hg19:

```{r annotation-hg19, eval=FALSE}
annotation_results <- ChIPSPannotation(
  chip_file = "r1881_hg19_test_peaks_ChIP.bed",
  chipsp_file = "ChIPSP_results.csv",
  species = "human",
  ref_genome = "hg19"
)
```

The corresponding genome annotation resource is:

```text
TxDb.Hsapiens.UCSC.hg19.knownGene
```

and human gene symbols are obtained using:

```text
org.Hs.eg.db
```

## Human hg38

For human data aligned to hg38:

```{r annotation-hg38, eval=FALSE}
annotation_results <- ChIPSPannotation(
  chip_file = "example_ChIP.bed",
  chipsp_file = "ChIPSP_results.csv",
  species = "human",
  ref_genome = "hg38"
)
```

The corresponding annotation resource is:

```text
TxDb.Hsapiens.UCSC.hg38.knownGene
```

## Mouse mm9

For mouse data aligned to mm9:

```{r annotation-mm9, eval=FALSE}
annotation_results <- ChIPSPannotation(
  chip_file = "example_ChIP.bed",
  chipsp_file = "ChIPSP_results.csv",
  species = "mouse",
  ref_genome = "mm9"
)
```

The corresponding resources are:

```text
TxDb.Mmusculus.UCSC.mm9.knownGene
org.Mm.eg.db
```

## Mouse mm10

For mouse data aligned to mm10:

```{r annotation-mm10, eval=FALSE}
annotation_results <- ChIPSPannotation(
  chip_file = "example_ChIP.bed",
  chipsp_file = "ChIPSP_results.csv",
  species = "mouse",
  ref_genome = "mm10"
)
```

The corresponding resources are:

```text
TxDb.Mmusculus.UCSC.mm10.knownGene
org.Mm.eg.db
```

## Annotation Distance

The default annotation window is:

```text
+/- 5000 bp
```

This is controlled by:

```r
binding_bp = 5000
```

For example:

```{r annotation-distance, eval=FALSE}
annotation_results <- ChIPSPannotation(
  chip_file = "r1881_hg19_test_peaks_ChIP.bed",
  chipsp_file = "ChIPSP_results.csv",
  species = "human",
  ref_genome = "hg19",
  binding_bp = 5000
)
```

Users may modify this distance according to the biological question.

## Annotation Outputs

By default, two files are generated:

```text
ChIP_anno_genes_upAdown_UCSC_Control.csv
ChIP_anno_genes_upAdown_UCSC_CHIPSP.csv
```

The first contains genes associated with conventional ChIP-seq peaks.

The second contains genes associated with ChIP-SP spatial regions.

# Step 5: Enrichr Analysis

`ChIPSPenrichment()` performs functional enrichment analysis using the output
from `ChIPSPannotation()`.

Because the default filenames produced by `ChIPSPannotation()` are also the
default input filenames expected by `ChIPSPenrichment()`, the standard analysis
can be run with:

```{r enrichment, eval=FALSE}
enrich_results <- ChIPSPenrichment()
```

The function constructs five gene sets:

```text
ChIP
ChIPSP
ChIPSP_only
ChIP_only
Shared
```

These represent:

- `ChIP`: genes associated with conventional ChIP-seq peaks
- `ChIPSP`: genes associated with ChIP-SP regions
- `ChIPSP_only`: genes detected only through ChIP-SP
- `ChIP_only`: genes detected only through conventional ChIP annotation
- `Shared`: genes detected by both approaches

## Default Enrichr Databases

The default analysis uses:

```{r enrichment-databases, eval=FALSE}
c(
  "GO_Biological_Process_2023",
  "GO_Cellular_Component_2023",
  "GO_Molecular_Function_2023",
  "KEGG_2021_Human",
  "Reactome_2022",
  "ChEA_2022",
  "MSigDB_Hallmark_2020"
)
```

The databases can be customized:

```{r enrichment-custom, eval=FALSE}
enrich_results <- ChIPSPenrichment(
  databases = c(
    "GO_Biological_Process_2023",
    "Reactome_2022",
    "ChEA_2022"
  )
)
```

The default database collection currently contains a human KEGG database.
Users performing mouse analyses should select Enrichr databases appropriate
for the organism and analysis.

## Enrichment Dotplots

For the generated dotplots:

- pathways are selected and ranked by nominal P value
- the smallest P value is displayed at the top
- dot color represents `-log10(P)`
- dot size represents the number of overlapping genes
- the x-axis represents gene ratio

Gene ratio is calculated as:

```text
number of overlapping genes
---------------------------
total genes in the pathway
```

The default significance threshold for displayed pathways is:

```text
P < 0.05
```

The default output directory is:

```text
Enrichr_dotplots/
```

# Complete Human hg19 Workflow

The following example shows the complete workflow for human hg19 data.

```{r complete-workflow, eval=FALSE}
library(ChIPSP)

# ------------------------------------------------------------
# Step 1
# Remove chromosome X and chromosome Y
# ------------------------------------------------------------

hic_files_clean <- removeXYChromosomes(
  path = getwd()
)


# ------------------------------------------------------------
# Step 2
# Merge Hi-C loops
# ------------------------------------------------------------

hic_df <- mergeHiCLoops(
  hic_files_clean
)


# ------------------------------------------------------------
# Step 3
# Run ChIP-SP
# ------------------------------------------------------------

chipsp_results <- chipSPLink(
  hic_df = hic_df
)


# ------------------------------------------------------------
# Step 4
# Gene annotation
# ------------------------------------------------------------

annotation_results <- ChIPSPannotation(
  chip_file = "r1881_hg19_test_peaks_ChIP.bed",
  chipsp_file = "ChIPSP_results.csv",
  species = "human",
  ref_genome = "hg19"
)


# ------------------------------------------------------------
# Step 5
# Enrichr analysis
# ------------------------------------------------------------

enrich_results <- ChIPSPenrichment()
```

# Complete Workflow Without Removing X and Y

The chromosome X/Y filtering step is optional.

If sex chromosomes should be retained:

```{r workflow-with-xy, eval=FALSE}
library(ChIPSP)

hic_files <- list.files(
  pattern = "HiC\\.(xls|xlsx|csv|tsv|txt|bed|tab)$",
  full.names = TRUE,
  ignore.case = TRUE
)

hic_df <- mergeHiCLoops(
  hic_files
)

chipsp_results <- chipSPLink(
  hic_df = hic_df
)

annotation_results <- ChIPSPannotation(
  chip_file = "r1881_hg19_test_peaks_ChIP.bed",
  chipsp_file = "ChIPSP_results.csv",
  species = "human",
  ref_genome = "hg19"
)

enrich_results <- ChIPSPenrichment()
```

# Main Output Files

A standard analysis produces several classes of output.

## Filtered Hi-C Files

When `removeXYChromosomes()` is used, filtered Hi-C files are generated with
chromosome X and chromosome Y interactions removed.

## ChIP-SP Spatial Results

The primary ChIP-SP result is:

```text
ChIPSP_results.csv
```

## Gene Annotation

The default annotation outputs are:

```text
ChIP_anno_genes_upAdown_UCSC_Control.csv
ChIP_anno_genes_upAdown_UCSC_CHIPSP.csv
```

## Enrichment Results

The enrichment workflow creates:

```text
Enrichr_dotplots/
```

This directory contains:

- gene lists
- gene-set summary
- individual Enrichr database result tables
- combined enrichment results
- PDF enrichment dotplots
- PNG enrichment dotplots

# Recommended Workflow Structure

The overall ChIP-SP analysis can be summarized as:

```text
Hi-C loop files
       |
       v
removeXYChromosomes()
       |
       v
mergeHiCLoops()
       |
       v
Merged Hi-C loop table
       |
       +-------------------+
       |                   |
       |              ChIP-seq peaks
       |                   |
       +---------+---------+
                 |
                 v
             chipSPLink()
                 |
                 v
        ChIPSP_results.csv
                 |
                 v
        ChIPSPannotation()
                 |
        +--------+--------+
        |                 |
        v                 v
Conventional ChIP     ChIP-SP genes
      genes
        |                 |
        +--------+--------+
                 |
                 v
        ChIPSPenrichment()
                 |
                 v
      Functional enrichment
```

# Important Considerations

ChIP-SP performance depends directly on the quality of the underlying ChIP-seq
and Hi-C data.

Users should consider:

- ChIP-seq peak quality
- ChIP-seq signal strength
- Hi-C sequencing depth
- Hi-C loop-calling quality
- Hi-C loop-calling resolution
- the number of detected Hi-C loops
- chromosome naming conventions
- compatibility of genome builds
- the gene annotation strategy used downstream

The ChIP-seq and Hi-C datasets must correspond to the same genome build before
spatial integration.

For example:

```text
Correct:
hg19 ChIP-seq + hg19 Hi-C + hg19 annotation

Incorrect:
hg19 ChIP-seq + hg38 Hi-C + hg19 annotation
```

If datasets use different genome builds, genomic coordinates should be
converted to a common reference genome before running ChIP-SP.

# Session Information

For reproducibility, session information can be recorded using:

```{r session-info}
sessionInfo()
```
